---
title: "Bates"
author: "April Wright"
date: '2022-07-21'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The first thing we need to do with this dataset is make sure the row labels and the tree tips are aligned. We'll use the package `treedata.table`, which automates many tree managemend functions for this.

```{r}
library(tidyverse)
library(treedata.table)
library(ape)
```

Our trait data need to be in a dataframe, and our trees must be read in as a `Multiphylo` data type.

```{r}

trees <- ape::read.nexus("trees/output.nex")
df <- read_delim("data/Bat Migration Phylogeny - All Species_Filled_woCite.tsv")
small_dat <- df %>%
  select(speciesName, migratory.sedentary)
sd <- as.data.frame(small_dat)
````

Then, the tree and data objects can be merged together for use with downstream comparative methods.

```{r}
td <- as.treedata.table(tree = trees, data = sd, name_column = "speciesName")
```

This will result in a few tips being pruned from the tree, as they do not have data.


Next, we will use `Revticulate` to run our scripts.

```{R}
library(Revticulate)
callRevFromTerminal("scripts/mcmc_ase_ERM.Rev")

```

Next, we will plot these states on the tree

```{r}
library(RevGadgets)
library(ggplot2)
CHARACTER <- "migration status"
NUM_STATES <- 2
i <- 1

while (i < 100){
STATE_LABELS <- c("0" = "Sedentary", "1" = "Migratory")

#treefile <- paste0("output/erm/migration",i,  "_ase_ERM.tree")
treefile <- paste0("output/freek/solitariness",i,  "_ase_freeK.tree")

ase <- processAncStates(treefile,
                        # Specify state labels.
                        # These numbers correspond to
                        # your input data file.
                        state_labels = STATE_LABELS)


plt <- plotAncStatesMAP(t = ase, 
                 node_color_as = "state_posterior",
                 node_color = c("orange", "blue"),
                 node_size_as = NULL,
                 node_size = c(1),
                 
                 tree_layout = "rectangular") + 
    # modify legend location using ggplot2
    labs("Posterior Probability of Sedentary")
    outfile <- paste0(treefile, ".png")
    ggsave(filename = outfile, plot = plt, dpi = "retina")
    i <- i + 1
}

```

```{r}

library(ggplot2)
CHARACTER <- "migration status"
NUM_STATES <- 2
i <- 1

while (i < 100){
STATE_LABELS <- c("0" = "Sedentary", "1" = "Migratory")

  treefile <- paste0("output/erm/migration",i,  "_ase_ERM.tree")
 # treefile <- paste0("output/freek/solitariness",i,  "_ase_freeK.tree")
  tree_sample <- readTrees(paths = treefile)
  single_tree <- tree_sample[[1]][[1]]

  single_tree@data$anc_state_1_pp <-as.double(single_tree@data$anc_state_1_pp)
  trait <- td$dat
  x <- tibble(label = as.phylo(single_tree)$tip.label, trait = trait)
  with_traits <- full_join(single_tree, x, by="label")
  p <- ggtree(with_traits, aes(color=anc_state_1_pp)) + scale_color_continuous(low ="red", high="darkblue") + geom_tippoint(data = 
with_traits@extraInfo)
    outfile <- paste0(treefile, "branchcolor", ".png")
    ggsave(filename = outfile, plot = plt, dpi = "retina")
    i <- i + 1
}

```


Now, let's do the free state estimation. This allows the forward and backward transition rates to vary.

```{R}
library(Revticulate)
callRevFromTerminal("scripts/mcmc_ase_freeK.Rev")
```

In order to have some baseline number of transitions, we will also do this for a consensus tree:

```{r}
consensus(trees, p = .5, check.labels = FALSE, rooted = FALSE)
td <- as.treedata.table(tree = single_tree, data = sd, name_column = "speciesName")
```

Then, rerun the RevScripts, taking as input the single consensus tree.


To extract the rates of back and forward transitions and summarize across files:

```{r}

library(tidyverse)

i <- 1
means <- c()

while ( i < 100){
  log <- paste0("output/erm/migration",i, "_ERM.log" )
  dat <- read_delim(log, delim ="\t")
  dat <- dat %>% 
    drop_na()
  means <- c(means, mean(dat$Posterior))

#  means <- c(means, mean(dat$`rate[1]`))
  i <- i+1
}
```

Next, we want to plot these data.

```{r}
df <- as.data.frame(means, means)
 plt <- ggplot(df, mapping=aes(x = means)) + geom_histogram() + ylab("Count") + xlab("Posterior") + ggtitle("One-Rate Model") + geom_vline(xintercept = -1602.888)
ggsave(plt, filename = "posterior_one-rate.pdf")
```

We also want to do this for the two-rate model. 
```{r}

library(tidyverse)

i <- 1
means_forward <- c()
means_backward <- c()

while ( i < 89){
  log <- paste0("output/freek/solitariness",i, "_freeK.log" )
  dat <- read_delim(log, delim ="\t")
  dat <- dat %>% 
    drop_na()
   means <- c(means, mean(dat$Posterior))
df <- as.data.frame(means)
#  means_forward <- c(means_forward, mean(dat$`rate[1]`))
 # means_backward <- c(means_backward, mean(dat$`rate[2]`))

  i <- i+1
}

 plt <- ggplot(df, mapping=aes(x = means)) + geom_histogram() + ylab("Count") + xlab("Posterior") + ggtitle("Two-Rate Model") + geom_vline(xintercept = -967.01)
ggsave(plt, filename = "posterior_two-rate.pdf")
```

Next, we want to plot these data.

```{r}
df_forward <- as.data.frame(means_forward)
colnames(df_forward) <- "means"

df_forward <- df_forward %>% 
  mutate(Rate = "Forward")
df_back <- as.data.frame(means_backward)
colnames(df_back) <- "means"
df_back <- df_back %>% 
  mutate(Rate = "Backward")  
df_rates <- rbind(df_forward, df_back)
plt <- ggplot(df_rates, mapping=aes(x = means, fill = Rate)) + geom_histogram(alpha = .75) + ylab("Count") + xlab("Rate of Transition") + ggtitle("Two-Rate Model") +theme_bw()
```

## Include tip colors: 
```{r}

library(ggplot2)
CHARACTER <- "migration status"
NUM_STATES <- 2
i <- 1

while (i < 100){
STATE_LABELS <- c("0" = "Sedentary", "1" = "Migratory")

  treefile <- paste0("output/erm/migration",i,  "_ase_ERM.tree")
 # treefile <- paste0("output/freek/solitariness",i,  "_ase_freeK.tree")
  tree_sample <- readTrees(paths = treefile)
  single_tree <- tree_sample[[1]][[1]]

  single_tree@data$anc_state_1_pp <-as.double(single_tree@data$anc_state_1_pp)
  trait <- td$dat
  x <- tibble(label = as.phylo(single_tree)$tip.label, trait = trait)
  with_traits <- full_join(single_tree, x, by="label")
  p <- ggtree(with_traits, aes(color=anc_state_1_pp)) %<+% df
  pp <- p +scale_color_continuous(low ="red", high="darkblue")

    outfile <- paste0(treefile, "branchcolor", ".png")
    ggsave(filename = outfile, plot = plt, dpi = "retina")
    i <- i + 1
}
